Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-01 tagless/tagless-final.tex
checking in more small edits a bit all over

\begin{abstract}
We have built a family of tagless embeddings (interpretations) for a
higher-order typed object
language in a typed metalanguage (current Haskell and ML), 
requiring neither dependent
types, nor generalized algebraic data types, nor a tag elimination 
postprocessing step. That has prevously considered an open problem
(impossible). 
Our family of evaluators (interpreters, embeddings?) contains 
an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
We take advantage of staging when available.

The main tool is to take functions seriously: we encode the
deBruijn or higher-order abstract syntax  of object terms 
not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
(Or: we expose the co-algebraic structure of the object language).
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), available with ML functors and
Haskell98 constructor classes. This lets us encode object terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.

@@

We focus on embedding a typed object language into a typed metalanguage.
Types are useful as usual because they document the programmer's
intentions and can be checked early by machine (see~\S\ref{encoding}).
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just for
evaluation of object programs but also for
compilation, partial evaluation, and other processing.

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions.
\begin{code}
let varZ (x,env)    = x
let varS vp (x,env) = vp env
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
These functions comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment.
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
The result has no tags---the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as |testf3| below gives a type error rather than
a run-time error.
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.
\ccshan{Need to cite Thiemann here and in related work?}

@@

Our contributions are
\begin{enumerate}
\item to evaluate, compile, and partially evaluate a typed object language
   in a typed metalanguage with a simple type system;
\item to check statically for well-typed and closed object programs;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CBN CPS
    transformation;
\item a clear way to extend the object language with more features;
\item and an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out to be harder than expected.
\end{enumerate}
\jacques{can we state fairly clearly what type system we need?  I mean,
Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
With our let-polymorphism trick for self-interpretation, what do we really
need?}

The structure of our paper is as follows. Section XXX ...
We finally conclude.

@@

We embed this language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics|. The class is so named because its interface gives the syntax for
the object language and its instances give the semantics.
The parameter |repr| to the class has the kind |* -> *|.
\begin{code}
class Symantics repr where

@@

Thus we represent an object expression in
OCaml as a functor from |Symantics|. This
is essentially the same as the constraint |Symantics repr =>| in the
Haskell embedding.

@@

The first interpreter evaluates the object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation.
\begin{code}
module R = struct

  (* absolutely no wrappers *)
  type ('c,'dv) repr = 'dv

@@

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R| module.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is compiled to no-op.
In both OCaml and Haskell, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
That is, the proposition below follows immediately from the soundness of
the metalanguage's type system.
\begin{prop}
If the term |e| encoded in the metalanguage has type~|t|,
then evaluating~|e| in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~|t|.
\end{prop}

@@

Our |Symantics| class includes only separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid the need for polymorphic lift, we turn to
a technique due to Asai \ccshan{cite?}: build a dynamic term alongside
every static term.

@@

When we come to the higher-order fragment of our language, however, we
stumble. Let us consider |lam f|. Here, |f| is a function from a
|P1 a| value to a |P1 b| value. The result of |lam f| must be either a
static value or a dynamic value -- but which? Unlike |add| or |if_|, 
|lam| cannot really examine its operand. The arguments of |add| are 
of type |P1 int|, which we can examine by pattern-matching and so 
determine if they are static or dynamic. The argument of |lam| is a 
function, which is not introspectible. One may only 
apply |f| to a particular static or dynamic value, and receive another 
value in return. Alas, as in the case |lam (\x -> mul x dynval)|, 
applying the same function to different static values may give 
either static or dynamic results (in the example above, applying 
the function to (int 0) gives the static value in return). One
solution is to be conservative and always use the |C| compiler to
compile |f| to the future-stage expression, which we can tag with
E1. But that defeats the purpose of partial evaluation as we may miss
many opportunities for static computations and simplifications, such
as multiplication by zero in the example above. A better solution is to
delay the binding-type-analysis (BTA) until the function |f| is
applied to a particular argument (that is, until the result of |lam f|
appears as the argument of |app|). That means, we have to find a way
to incorporate |f| as it is into the P1 data structure. We would like
to write that when |t| is a particular type |a->b|, that data type
should be 
\begin{code}
data P1 (a->b) = S1 (R (P1 a -> P1 b)) (C (a->b)) 
               | E1 (C (a->b))
\end{code}
That is, we need a non-parameteric data type, something like a
typecase (ref? Gibbons?).

The problem we encountered here is quite akin to typed CPS
transform. When transforming abstractions, we come across
non-functorialness. The connection with CPS is not an accident, as we
shall see below (in Section XXX).

@@

Two common ways of providing typecase in Haskell is by using either
generalized algebraic data types (GADT) or type-class functions
(implemented via typeclasses with functional dependencies). These
methods are equivalent; here we show the GADT; please see |incope1.hs|
in the accompanying source code for the latter.

We introduce the following GADT
\begin{code}
data P t where
    VI :: Int  -> P Int
    VB :: Bool -> P Bool
    VF :: (P a -> P b) -> P (a->b)
    E  :: C t -> P t
\end{code}
with four data constructors. The constructor |VF| stands out as its
operand is of the type |P a -> P b| rather than |a -> b| as one may
have expected if |P t| were parametric in |t|. As before, we define a
function to extract the future-stage computation from the |P t| value:
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}

We use the method |lam| of the |C| intrepreter to compile the argument
of |lam| into the code value. We may now make |P| the instance of
Symantics and thus implement the partial evaluator (the
implementations of |mul|, |leq|, |if_| and |fix| are elided).
\begin{code}
instance Symantics P where
    int x  = VI x
    bool b = VB b

@@

The implementations of |int|, |bool|, and |add| are similar to the
above. The interpretation of |lam f| merely incorporates the
HOAS function |f| into the result. At any point we can convert it to a
code value, if we must. But delaying such a conversion, we get the
opportunity to apply |f| to a concrete argument. When we evaluating
|app ef ea|, we check to see of the first argument is so-delayed
lambda-expression. If it is, we apply the incorporated |f| to the
concrete argument |ea|, giving us a chance to perform the static
computations (we see the example in the next subsection). If |ef| is a
dynamic value, we residualize.

Although GADT solve our problem, the solution is not quite
satisfactory. First of all, it cannot be ported to MetaOCaml as GADTs
are not supported in OCaml. Second, the tagging problem reappeared, in
the implementation of |app| above. The datatype |P t| has four data
constructors; the pattern-matching in |app| however used only two of
them, |VF| and |E|. It is obvious, one may say, that the constructors
|VI| and |VB| just cannot occur, because the corresponding values
cannot have the arrow type. What is obvious to us is not obvious to
the compiler. The case matching above does look \emph{partial}. The
point may loop tenuous -- and yet it is precisely the main point of
tag elimination. In a typed tagged interpreter there are many
pattern-matching forms that looks partial but are total in
reality. The goal is to make this totality \emph{syntactically}
apparent.

\subsection{The `final' solution}
\label{S:PE-solution}
Let us re-examone the problem in Section~\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = S (repr_pe t) (C t) | E (C t)
\end{code}
where |repr_pe| is a type function, defined inductively
% inductively because P below depends on repr_pe
as 
\begin{code}
repr_pe Int    = Int
repr_pe Bool   = Bool
repr_pe (a->b) = P a -> P b
\end{code}
Although we can use typeclasses to define that type function
in Haskell, that is not portable to MetaOCaml. We notice however that
the type case alternatives above are all associated with the existing
methods of |Symantics| and so the case analysis is already present. 
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml for
portability, we recall from Section~\ref{encoding} that |repr| type
had two arguments |type ('c,'dv) repr|. We add an extra argument,
|'sv|, which is the result of applying |repr_pe| function above to the
type |'dv|. Figure~\ref{fig:ocaml} shows the new signature |Symantics|.
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

@@

The interpreters |R|, |L| and |C| above only relied on the last
argument of the |repr| type, which is unchanged. And so is their
implementation; the newly added argument |'sv| will be phantom.
For example, we would now write the interpreter |C| as
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
When writing the partial evaluator however, we make use of the
type argument |'sv|. Here's the corresponding instance of the
Symantics:

@@

The type |repr| here literally expresses the type equation at the
beginning of this section. The function |abstr|, as in the previous
subsection, extracts the future-stage computation from the result of
PE. The function |pdyn| does the converse. As in
section~\ref{S:PE-problem}, we build future-stage computations as we
build present-stage ones, if any. Therefore, we avoid the need for the
polymorphic `lift' function. The implementations of
|add| and |mul| are a bit more sophisticated, taking advantage of the
implicit monoid structure. The function |buildsimp| checks to see if
its two arguments |e1| and |e2| are static expressions. If so, the
corresponding static computation (with the help of the intrepreter
|R|) is performed. The function also checks if one argument is static
and is the neutral element of the monoid structure 
\jacques{Comment on the various nice generic things in the implementation,
which are a little silly in this small version, but would really pay off
in something bigger.  See the abstract interpretation hard at work!
See the monoid-driven simplifications!}
\oleg{I'm inclined to ditch buildsimp. We miss an opportunity here:
  when multiplying anything by 0, the result is zero!}

@@

\subsection{Call-by-name CPS interpreters}

@@

>>>>

\begin{abstract}
We have built a family of tagless embeddings (interpretations) for a
higher-order typed object
language in a typed metalanguage (current Haskell and ML), 
requiring neither dependent
types, nor generalized algebraic data types, nor a tag elimination 
postprocessing step. That has prevously considered an open problem
(i.e. impossible). 
Our family of evaluators contains 
an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
We take advantage of staging when available.

The main tool is to take functions seriously: we encode the
de Bruijn or higher-order abstract syntax  of object terms 
not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
(Or: we expose the co-algebraic structure of the object language).
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), available with ML functors and
Haskell98 constructor classes. This lets us encode object terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.

@@

We focus on embedding a typed object language into a typed metalanguage.
Types are useful as usual because they document the programmer's
intentions and can be checked early (see~\S\ref{encoding}).
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just for
evaluation of object programs but also for
compilation, partial evaluation, and other processing.

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions.
\jacques{but now this smaller language does not match anymore what
we wrote earlier.  I think it is better to use a small language here,
but i don't quite know how to say so}
\begin{code}
let varZ (x,env)    = x
let varS vp (x,env) = vp env
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
These functions comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment.
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
The result has no tags---the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as |testf3| below gives a type error rather than
a run-time error.
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.  See \citep{Gluck-jones-optimality}
for deeper technical reasons which inevitably lead to these performance
penalties.

@@

Our contributions are
\begin{enumerate}
\item to evaluate, compile, and partially evaluate a typed object language
   in a typed metalanguage with a simple type system;
\item to check statically for well-typed and closed object programs;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CBN CPS
    transformation;
\item a clear way to extend the object language with more features;
\item and an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out to be harder than expected.
\end{enumerate}

The structure of our paper is as follows. Section XXX ...
We finally conclude.
\jacques{TODO}

@@

We embed this language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics|. The class is so named because its interface gives the syntax for
the object language and its instances give the semantics.
The parameter |repr| to the class has kind |* -> *|.
\begin{code}
class Symantics repr where

@@

Thus we represent an object expression in
OCaml as a functor from |Symantics| to an appropriate semantic domain. This
is essentially the same as the constraint |Symantics repr =>| in the
Haskell embedding.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation.
\begin{code}
module R = struct
  (* absolutely no wrappers *)
  type ('c,'dv) repr = 'dv

@@

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R|
module.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is compiled away.
In both OCaml and Haskell, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
That is, the proposition below follows immediately from the soundness of
the metalanguage's type system.
\begin{prop}
If the term |e| encoded in the metalanguage has type~|t|,
then evaluating~|e| in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~|t|.
\end{prop}

@@

Our |Symantics| class includes only separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid the need for polymorphic lift, we turn to
a technique due to Asai \ccshan{cite?}\jacques{yes}: build a dynamic term
alongside every static term.

@@

When we come to the higher-order fragment of our language, however, we
stumble. Let us consider |lam f|. Here, |f| is a function from a
|P1 a| value to a |P1 b| value. The result of |lam f| must be either a
static value or a dynamic value -- but which? Unlike |add| or |if_|, 
|lam| cannot really examine its operand. The arguments of |add| are 
of type |P1 int|, which we can examine by pattern-matching and so 
determine if they are static or dynamic. The argument of |lam| is a 
function, whose structure cannot be observed. One may only 
apply |f| to a particular static or dynamic value, and receive another 
value in return. Alas, as in the case |lam (\x -> mul x dynval)|, 
applying the same function to different static values may give 
either static or dynamic results (in the example above, applying 
the function to (int 0) gives a static value in return). One
solution is to be conservative and always use the |C| compiler to
compile |f| to a future-stage expression, which we can tag with
E1. But this defeats the purpose of partial evaluation as we may miss
many opportunities for static computations and simplifications, such
as multiplication by zero in the example above. A better solution is to
delay the binding-type-analysis (BTA) until the function |f| is
applied to a particular argument (that is, until the result of |lam f|
appears as the argument of |app|). In other words, we have to find a way
to incorporate |f| as it is into the P1 data structure. We would like
to write that when |t| is a particular type |a->b|, this data type
should be 
\begin{code}
data P1 (a->b) = S1 (R (P1 a -> P1 b)) (C (a->b)) 
               | E1 (C (a->b))
\end{code}
That is, we need a non-parametric data type, something akin to
type-indexed functions and type-indexed types, as used in 
\citet{Oliveira2005:TypeCase}, where it is dubbed the \emph{typecase}
design pattern.

The problem we encountered here is quite akin to the typed CPS
transform. When transforming abstractions, we come across
non-functorialness. The connection with CPS is not an accident, as we
shall see in~\S\ref{S:CPS}.

@@

Two common ways of providing typecase in Haskell is by using either
generalized algebraic data types (GADT) or type-class functions
(implemented via typeclasses with functional dependencies)
\citet{Oliveira2005:TypeCase}.  These
methods are equivalent, and here we show the GADT; please see |incope1.hs|
in the accompanying source code for the latter.

We introduce the following GADT
\begin{code}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a->b)
  E  :: C t -> P t
\end{code}
with four data constructors. The constructor |VF| stands out as its
operand has type |P a -> P b| rather than |a -> b| as one may
have expected if |P t| were parametric in |t|. As before, we define a
function to extract the future-stage computation from the |P t| value:
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}

We use the method |lam| of the |C| intrepreter to compile the argument
of |lam| into a code value. We may now make |P| an instance of
Symantics and implement the partial evaluator (the
implementations of |mul|, |leq|, |if_| and |fix| are elided) as follows.
\begin{code}
instance Symantics P where
int x  = VI x
bool b = VB b

@@

The implementations of |int|, |bool|, and |add| are similar to the
above. The interpretation of |lam f| merely incorporates the
HOAS function |f| into the result. At any point we can convert it to a
code value, if we must. By delaying such a conversion, we get the
opportunity to apply |f| to a concrete argument. When evaluating
|app ef ea|, we check to see of the first argument is such a delayed
lambda-expression. If it is, we apply the incorporated |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix| in the next subsection). If |ef| is a
dynamic value, we residualize.

Although GADTs solve our problem, the solution is not quite
satisfactory. First of all, it cannot be ported to MetaOCaml as GADTs
are not supported there. Second, the tagging problem reappeared, in
the implementation of |app| above. The datatype |P t| has four data
constructors; pattern-matching in |app| however uses only two of
them, |VF| and |E|. It is obvious, one may say, that the constructors
|VI| and |VB| just cannot occur, because the corresponding values
cannot be function types. What is obvious to us is not obvious to
the compiler. The case matching above looks \emph{partial}.  While this
may seem like a minor point, 
this is precisely the main point of
tag elimination. In a typed tagged interpreter there are many
pattern-matching forms that looks partial but are total in
reality. The goal is to make this totality \emph{syntactically}
apparent.

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in Section~\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = S (repr_pe t) (C t) | E (C t)
\end{code}
where |repr_pe| is a type function, defined inductively
% inductively because P below depends on repr_pe
as 
\begin{code}
repr_pe Int    = Int
repr_pe Bool   = Bool
repr_pe (a->b) = P a -> P b
\end{code}
Although we can use typeclasses to define this type function
in Haskell, that is not portable to MetaOCaml. We notice however that
the type case alternatives above are all associated with existing
methods of |Symantics|, and so the case analysis is already present. 
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml for
portability, we recall from Section~\ref{encoding} that the |repr| type
had two arguments |type ('c,'dv) repr|. We add an extra argument,
|'sv|, which is the result of applying the |repr_pe| function above to the
type |'dv|. Figure~\ref{fig:ocaml} shows the new signature |Symantics|.
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

@@

The interpreters |R|, |L| and |C| above only relied on the last
argument of the |repr| type, which is unchanged. And so is their
implementation; the newly added argument |'sv| will be phantom.
For example, we would now write the interpreter |C| as
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
When writing the partial evaluator however, we make use of the
type argument |'sv|. Here is the corresponding instance of
|Symantics|:

@@

The type |repr| here literally expresses the type equation at the
beginning of this section. The function |abstr|, as in the previous
subsection, extracts the future-stage computation from the result of
partial evaluation.  The function |pdyn| does the converse. As in
section~\ref{S:PE-problem}, we build future-stage computations at the
same time as we
build present-stage ones, if any. Therefore, we avoid the need for a
polymorphic `lift' function. The implementations of
|add| and |mul| are a bit more sophisticated, taking advantage 
respectively of the implicit monoid and ring structure of the 
underlying domain.
The function |buildsimp| checks to see if
its two arguments |e1| and |e2| are static expressions. If so, the
corresponding static computation (with the help of the intrepreter
|R|) is performed. The function also checks if one argument is static
and is the neutral element of the monoid structure 
\jacques{Comment on the various nice generic things in the implementation,
which are a little silly in this small version, but would really pay off
in something bigger.  See the abstract interpretation hard at work!
See the monoid-driven simplifications!}
\oleg{I'm inclined to ditch buildsimp. We miss an opportunity here:
  when multiplying anything by 0, the result is zero!}

@@

\subsection{Call-by-name CPS interpreters}\label{S:CPS}

@@

Thiemann, cogen in six lines (for the interpreter/compiler work)?
